---
title: "Calculating word embeddings of random wikipedia articles using Singular Value Decomposition"
format: html
---

## Summary
While going back to work on the underlying model of this reccommendation system, I realized I did not have a perfect grasp of the theory behind the creation of word embeddings. This workbook is an attempt to rectify this. In it, I web-scraped random wikipedia articles, create a word corpus and perform Singular Value Decomposition on them, sort of like practice. The results are quite impressive.

## Data Collection
Wikipedia provides a url that spits out random articles. Unfortunately, this cannot be seeded either in R on in wikipedia itself (as far as I know). Therefore, every iteration of this practice session would give differing results. However, I beleive the uderlying insights would be the same.

```{r}
# --- Loading necessary libraries
library(rvest)
library(dplyr)
library(stringr)
library(tidytext)
library(tidyr)
library(SnowballC)
```

Now to create a function to pull out the words of random wikipedia articles

```{r}
# Creating function to collect text from random wikipedia pages
get.wiki.corpus <-
function(documents){
  i = 1

  output <- tibble()

  while (i <= documents) {
    page = read_html("https://en.wikipedia.org/wiki/Special:Random")

    title = page %>%
      html_nodes("h1") %>%
      html_text %>%
      paste0(collapse = "|")

    text <-
    page %>%
      html_nodes("p") %>%
      html_text %>%
      paste0(collapse = ". ")

    output <- output %>%
      bind_rows(
        tibble(
          title = title,
          text = text 
        )
      )
    
    i = i + 1

  }
  
  return(output)
}
wiki.corpus <- get.wiki.corpus(documents = 2500)
wiki.corpus
```

## Data Cleaning
Due to scope of excersise, minimal cleaning would be done. Only the removal of paragraph annotation and wikipedia reference boxes.

```{r}
# adding row number and perform minimal cleaning
wiki.corpus <- wiki.corpus %>%
  mutate(
    document = row_number(),
    text = str_remove_all(text, "\\n|\\[\\d+\\]")
  ) %>%
  select(document, title, text)
wiki.corpus
```

## Data Formating
The arrangement for the training of embeddings would be documents in rows, words in columns and term frequency - inverse document frequency as values. This would be the end goal of data formating. Firstly, I calculate the tf_idf of all words per document.

```{r}
# making tfidf values
wiki.tfidf <-
wiki.corpus %>%
  select(document, text) %>%
  unnest_tokens(word, text, token = "words") %>%
  group_by(document) %>%
  count(word) %>%
  filter(word != "") %>%
  bind_tf_idf(term = word, document = document, n = n)
wiki.tfidf
```

Then, I cast into a sparse matrix for smaller size and efficiency.

```{r}
# create a sparse matrix
wiki.sparse <-
wiki.tfidf %>%
  group_by(document) %>%
  cast_sparse(row = document, column = word, value = tf_idf)
wiki.sparse[1:10, 1:10]
```

## Mathematical Background and Computation

The mathematical methodlogy being used is Singular Vaue Decomposition, which is essentially a way to factorize or break down a matrix into three matrixes. The formula is given below:

$$
A = U \Sigma V^T
$$

Manually breaking them down would required finding the eigen values for $ AA^T $ in the case of $ U $ (left singular matrix) or  $ A^TA $ in the case of $ V $ (right singular matrix). Luckily, base R gives us a method for tackling this: 

```{r}
# Declare number of dimensions to be trained for
dimension_num <- 400

# Use base function to perform Singular Value Decomposition
wiki.svd <- svd(wiki.sparse, nu = dimension_num, nv = dimension_num)
str(wiki.svd)
```

Note that the dimensions are restricted to 400 to avoid computation of all dimensions.

The explaination for the results of Singular Value Decomposition is pretty straight forward. The multiplication of $ U\Sigma $ gives rise to the projection of the documents into the feature space. This could be used directly for prediction. On the other hand, $ V\Sigma $ are the individual word embeddings in the same feature space. They can be used to break down a sentence into its document embedding.

Before we continue, lets make a function that calculates the cosine of the angle between two vectors. Angles that are close to each other would have values closer to 1, other angles with values closer to -1 show great dismilarity and 0 can be infered as being neutral.

```{r}
cosine_similarity <- function(token, data, target_variable = "document"){
  data_matrix = as.matrix(data %>% select_if(!str_detect(colnames(.), target_variable)))

  rownames(data_matrix) = data[[target_variable]]
  colnames(data_matrix) = paste0("dim_", 1:ncol(data_matrix))

  token_matrix = matrix(data_matrix[rownames(data_matrix) == token,], nrow = 1)[rep(1, nrow(data_matrix)),]

  result = rowSums(data_matrix * token_matrix) / (sqrt(rowSums(data_matrix ^ 2)) * (sqrt(rowSums(token_matrix ^ 2))))

  output = tibble(
    document = names(result),
    cosine = result
  )

  return(output)

}
```

## Results and Review

Now, lets make our document embedding.

```{r}
document.embeddings <- (wiki.svd$u %*% diag(wiki.svd$d[1:dimension_num])) %>%
  as_tibble() %>%
    mutate(
      document = wiki.corpus$title
    ) %>%
    select(document, everything())
document.embeddings
```

If prediction was the goal of this project, this would be the training set.

Since this represents the position of various wikipedia articles in the trained vector stage, lets see which articles are close to each other.

```{r}
cosine_similarity(token = document.embeddings$document[1], data = document.embeddings) %>%
  arrange(-cosine) %>%
  print(n = 20)
```

Because of the random nature of the wikipedia url response, i cannot make definite analysis of the results. However, the first three or four articles should prove similar.

Now lets make a word embedding corpus.

```{r}
word.embeddings <- (wiki.svd$v %*% diag(wiki.svd$d[1:dimension_num])) %>%
  as_tibble() %>%
    mutate(
      word = colnames(wiki.sparse)
    ) %>%
    select(word, everything())
word.embeddings
```

This repressents every word that exist in the response. It represents every word in `dimension_num` dimensions, putting words that are similar in closer proximity to each other. Lets use the word "actor" since its quite popular in wikipedia and see which word is close to it

```{r}
cosine_similarity(token = "actor", data = word.embeddings, target_variable = "word") %>%
  arrange(-cosine) %>%
  print(n = 20)
```

Again not sure what the result will be but i am willing to bet that words related to movies would be represented. Related words you could see would include movie studios, movie jobs such as "director" or film related adjectives.

$$
\text{Success!}
$$

Now to apply to actual project.